---
title: "2025 Lab 1 Notes"
author: "Overview of Quantitative Methods"
output: 
  pdf_document: default
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---

# R basics

## Arithmetic, variables, vectors

You can use R as a glorified calculator for basic arithmetic:
```{r, echo = T}
2+3
```
you can save results of the operation to variable:
```{r, echo = T}
a <- 2+3
```
The operator `<-` assigns value to variable. You can also use it left to right (sometimes, it is useful):

```{r}
3+2 -> a
a = 2+3
```
You can also use `=` as most normal programming languages do, but it less flexible in R (i.e., you can't assign values with `=` within functions).

Now, you can call `a` and get its value:
```{r}
print(a)
```
There are different data types with their own properties and methods: 

- int, real, double etc. for numbers; 

- logic (TRUE/FALSE or just T/F);

- string ("hello world", "Password123, "42");

- date ("2024/01/01", "11/12/1994");

```{r}
class(a)
#?class
#help(class) #very useful commands
x <- "5"
class(x)
is.numeric(x)
is.data.frame(a)
is.logical(is.data.frame(a))
rm(x) #that's how you clear object from memory
#rm(list=ls()) #trivia time: what does this line?
```

You can assign the value of more complex functions and make operations with variables:
```{r, echo = T}
b <- 5*4-7/2+3^2 #more calculation
c <- pi^3 #R knows some constants
b-a
a+c
sum(a,b,c) #also there are built-in arithmetic functions
sqrt(b*a) #and more functions
log(a^c)
```
Note that when I type `#` in my code, R ignores everything following the `#`. This is called leaving a comment,
and it helps you annotate what your code is doing. Please comment your code! It will save your future self a
lot of hassle when you’re trying to figure out what two-months-ago-you was thinking.

You can also have lists of numbers called vectors, which is handy. You can make a vector from scratch using a
function called `c()` (for this reason, try not to name any variables c), but usually we’ll be loading real data
as vectors.
```{r}
# make some vectors
v1 <- c(1, 2, b)
v2 <- c(4, a, 6)
# add them together (if they're the same length)
v1 + v2
```
You can select values from the vector by indexing, such as
```{r}
v2[1] # get the first value in v2
```
__Unlike normal programming languages, R indexes vectors from 1, not 0!__
```{r}
v2[1:2] # get the first through the second value in v2
v2[-1] # unlike in Python, this is not the last element of v2, but v2 without the first element
v2[length(v2)] #this gives the last element
```
There are many built-in statistics for vectors (generally speaking, most functions work with vectors which is nice and also speeds up things):
```{r}
mean(v1)
sd(v2)
sum(v1)
median(v2)
length(v1) #size
cor.test(v1,v2) #correlation
t.test(v1,v2) #t-test
```
You can run logical operations on variables and vectors:
```{r}
b-a == 11
v1 != 2
v2 > 4
v1 <= 5
```

If you have a logical/boolean/true-or-false vector like you get from v1 >= 2, you can use that to index
another vector.
```{r}
# get only value of v1 that are greater than or equal to 2
idx <- v1 >= 2
v1[idx]
```
## Programming language

We can write loops (which are relatively ineffective in R) and functions (sometimes they are very useful!):
```{r}
for (i in v1) {
  print(i^2)
}

j <- 1
while (j < 3) {
  print(v2[j])
  j <- j + 1
}

pair.sum <- function(x, y) {
  out <- x + y
  return (out)
}
pair.sum(25, 4444)
```

## Distributions

R has a ton of built-in functions for statistical distributions:

```{r}
set.seed(1234) #save randomizer value for reproduction

#all distributions have the same style of syntaxis

#random draws from normal distribution with given parameters
rnorm(5, mean = 0, sd = 20) 

#Computes the probability density of a normal distribution at a specific value
dnorm(0.5, mean = 0, sd = 1)

#Cumulative density of normal distribution with given parameters
pnorm(7, mean = 5, sd = 10)

#Computes the quantile (or inverse cumulative probability) of a normal distribution
qnorm(0.27, mean = 10, sd = 25)
# calculates the value x such that the probability of the variable being less than or equal x is 0.27

x = seq(-5,5, 0.1)
dx = dnorm(x, mean = 0, sd = 1)
plot(x,dx)
px = pnorm(x, mean = 0, sd = 1)
plot(x,px)
x = seq(0,1, 0.05)
qx = dnorm(x, mean = 0, sd = 1)
plot(x,qx)

#Syntax works similarly with other distributions
#like uniform distribution

#Generates random numbers from a uniform distribution
runif(10, min = -1, max = 1) 

# Computes the probability density of a uniform distribution at a given value
dunif(0.2, min = -1, max = 1)

# Computes the cumulative probability (area under the curve) up to a given value
punif(0, min = -1, max = 1)

# Computes the quantile (inverse cumulative probability) of a uniform distribution
qunif(0.75, min = -1, max = 1)

#or Poisson
rpois(4, lambda = 0.5)
dpois(5, lambda = 2)
ppois(0.67, lambda = 0.67)
qpois(0.8, lambda = 10)

#etc.

```

# Data frames

Most often though we have more than one dimension in our data, and store the collection of vectors in data.frame:
```{r}
#create data frame with columns `id` and `height`
df <- data.frame(id = 1:10,
                 height = rnorm(10, mean = 5.7, sd = 1.5)) 
weight <- rnorm(10, mean = 160, sd = 20)
sex <- rep("m", 10) #repeat "m" 10 times
#create data frame from vectors weight and sex
df1 <- data.frame(weight, sex)

#that's also why data.frame columns are basically vectors:
df$height
df$height[2:4]

df <- cbind(df, df1) #bind columns by row
```


1. cbind() (Column Bind)

What it does: Combines two or more objects (e.g., vectors, matrices, data frames) by binding them as columns.

Use case: When the objects have the same number of rows and you want to add them side by side as new columns.

Output: A matrix or data frame, depending on the input types.


2. rbind() (Row Bind)

What it does: Combines two or more objects (e.g., vectors, matrices, data frames) by binding them as rows.

Use case: When the objects have the same number of columns (or column names match, for data frames) and you want to stack them vertically as new rows.

Output: A matrix or data frame, depending on the input types.


3. merge()

What it does: Combines two data frames based on one or more common columns or row names.

Use case: When you want to perform database-like operations, such as joining two data frames by a key column (e.g., primary/foreign key relationship).

Output: A data frame containing matched rows, with options for different types of joins (e.g., inner join, outer join).

```{r}
#you can index data.frames by row (note the comma!):
df[2,]
df[2:3,]

#or column
df[2] #df[,2]
df[2:3]

df['height']
df[, c("height", "weight")]
df$height

class(df['height'])
class(df$height)

#or both
df[2:3,2:3]

#or filter by condition(s)
df[df$height > 6.0,]
df[df$height > 6.0 & df$weight < 150,]
df[df$weight > 150,]$id

#because loops are stupid in R we apply* functions to vectors instead
#*(there are different kinds of apply):
sapply(df, is.numeric)

df2 <- data.frame(id = 11:20,
                 weight = rnorm(10, mean = 150, sd = 20))
df2$heightt <- rnorm(10, mean = 5.5, sd = 1.5) #you can also add columns like this
df2$sex <- "f" #or like this

names(df2) 
names(df2)[3] <- "height" #one way to rename columns

df3 <- rbind(df,df2) #bind rows by column

df3$id <- as.factor(df3$id) #change variable type
df3$female <- ifelse(df3$sex == "f", T, F)
df3$male <- as.numeric(df3$sex == "m")
df3$sex <- ifelse(df3$sex == "m", 'male', 'female') #or recode it

head(df3,3)
tail(df3,3)
```

Next, we can read and write the data from disk. We can do it manually or in script (but we need to check the working directory!). R can read virtually any type of table or text data with the right function (.xlsx, .xls, .dta, .sas, etc.).
```{r}
getwd() #get the working directory
#old_wd <- getwd()
#new_wd <- paste(new_wd)
#setwd() #you can paste the new working directory here

write.csv(df3, "data1.csv")
df <- read.csv("data1.csv")[,-1] #default csv writer/reader adds annoying own index column

str(df)
dim(df) #gives both the number of columns and rows

unique(df$sex)
table(df$sex)
table(df$sex, df$height >= 6.0)
aggregate(height ~ sex, data = df, FUN = mean, na.rm = TRUE)
```
R has some basic built-in plots:
```{r}
hist(df$weight)
boxplot(df$height~df$sex)
plot(df$height~df$weight)
```

# R markdown

Probably just inb-class demonstration and tutorial [link](https://bookdown.org/yihui/rmarkdown/notebook.html)?

```{r}
#install.packages(knitr)
#install.packages(tinytex)
```




