---
title: "Lab 2 Notes"
author: "Overview of Quantitative Methods"
output: 
  pdf_document: default
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---

# Data manipulation & Tidyverse

We can make our R experience much more pleasant with user-made libraries:
```{r}
#install.packages("readxl")
#install.packages("tidyverse")
library(readxl)
#(this is considered bad taste btw, and it's better to load only the tidyverse packages you need)
library(tidyverse)
```

We will work with the data from the [American coffee preference test](https://www.youtube.com/watch?v=bMOOQfeloH0) this time.
```{r}
gact <- read_xlsx("GACTT_RESULTS_ANONYMIZED_LAB2.xlsx")
gact_codebook <- read_xlsx("GACTT_RESULTS_ANONYMIZED_LAB2.xlsx", 
                           sheet = "codebook") #to read particular xlsx sheet
head(gact_codebook)
glimpse(gact)
```
Tidyverse is the set of libraries which make R lot cooler. For example, it has the pipe `%>%` which applies function to object:
```{r}
gact %>% head(3)
head(gact, 3) #equivalent
```
Tidyverse also introduces a ton of functions for data manipulation (with pipe we don't change the original object):
```{r}
select(gact, submission_id)
gact %>% select(submission_id)
gact %>% rename(submission_ID = submission_id)

gact %>% filter(age == "18-24 years old")
gact %>% filter(age == "18-24 years old" & cups >= 3.0) #what's the problem here?

table(gact$cups) #`cups` has weird categories while substantially it is a number
#how to recode strings:
gact$cups_number <- gact$cups %>% recode("Less than 1" = '0', "More than 4" = '5')
#time to change type:
gact %>% mutate(cups_num = as.numeric(cups_number),
                cups_norm = cups_num - mean(cups_num, na.rm = T)) #remove NA = TRUE
gact$cups_num <- as.numeric(gact$cups_number)

gact %>% arrange(cups_num, desc = T)
```
With pipe we can also beautifully chain functions (you can think of the syntax as and SQL query)
```{r}
#let's also create the numeric column for age
table(gact$age)

gact %>% 
  select(submission_id, cups_num, age) %>% 
  filter(cups_num >= 3) %>% 
  mutate(age_num = str_remove(age, " years old"))

#equivalent to, but more readable!
mutate(filter(select(gact, submission_id, cups_num, age), cups_num >= 3), 
       age_num = str_remove(age, " years old"))

gact$age_num <- gact$age %>% 
  recode("<18 years old" = 18, 
         ">65 years old" = 65, 
         "18-24 years old" = 21, 
         "25-34 years old" = 30, 
         "35-44 years old" = 40, 
         "45-54 years old" = 50, 
         "55-64 years old" = 60)
```
Also, we can make summaries:
```{r}
gact %>% 
  summarize(mean_cups = mean(cups_num),
            mean_cups_narm = mean(cups_num, na.rm = T),
            median_cups = median(cups_num, na.rm = T))

gact %>% 
  group_by(gender) %>% #same, but for each gender
  summarize(mean_cups = mean(cups_num, na.rm = T),
            median_cups = median(cups_num, na.rm = T),
            n_obs = n())

gact %>% 
  select(gender, age, cups_num) %>% 
  filter(age %in% c("<18 years old", "18-24 years old", "25-34 years old")) %>% 
  group_by(gender, age) %>% #we can group by many columns
  summarize(mean_cups = mean(cups_num, na.rm = T),
            median_cups = median(cups_num, na.rm = T),
            n_obs = n())
```
Now to some serious things. Question "where do you drink coffee" (variable `where` in our data) is multiple-choice and the result is total mess, let's make nice dummies from it.
```{r}
gact %>% select(where) %>% table() %>% head()
gact %>% 
  drop_na(where) %>% #we don't need missing observations for this
  select(submission_id, where) %>% 
  separate_rows(where, sep = ",\\s*") %>% #that's regex and it's black magic
  # we flattened our data so each submission now has as many rows as there were choices 
  mutate(dummy = 1)
gact %>% 
  #drop_na(where) %>% #this time let's not drop the missing data
  select(submission_id, where) %>% 
  separate_rows(where, sep = ",\\s*") %>%
  mutate(dummy = 1) %>%
  #pivot_* is a VERY useful function for transforming your data from long to wide format or vice versa
  pivot_wider(names_from = where, values_from = dummy, values_fill = 0) 
```
We can of course just mutate the original dataset, but let's also practice merging data:
```{r}
gact_where <- gact %>% 
  select(submission_id, where) %>% 
  separate_rows(where, sep = ",\\s*") %>%
  mutate(dummy = 1) %>%
  pivot_wider(names_from = where, values_from = dummy, values_fill = 0) %>% 
  select(-`NA`) %>% #drop `NA` column
  rename("where_cafe" = "At a cafe", #rename to get rid of whitespaces
         "where_office" = "At the office",
         "where_home" = "At home",
         "where_go" = "On the go",
         "where_other" = "None of these")


gact_new <- merge(gact, gact_where, by='submission_id')

# same thing
gact <- left_join(gact, #identical to SQL left join
                  gact_where, #to each row in gact add data from gact_where
                  by = "submission_id") #matching by `submission_id` value

```

## GGplot
Also tidyverse is home to the superior graphic device called ggplot2. It's using the following basic syntax:
```{r, eval = F}
ggplot(df, #data  
       aes(x, y) #axis 
        )+ 
  geom_*()+ 
  coord_*()+ #usually omitted 
  theme_*()
```

Let's start with some quick examples
```{R}
ggplot(gact, aes(x = cups_num))+
  geom_histogram(bins = 6)

ggplot(gact, aes(x = age_num))+
  geom_histogram(bins = 5)+
  coord_flip()

ggplot(gact, aes(x = gender, y = cups_num))+
  geom_boxplot()+
  theme_minimal() #nice minimalist theme

ggplot(gact, aes(x = age_num, y = cups_num, color = gender))+
  geom_point()+ #scatter plot
  theme_minimal()

ggplot(gact, aes(x = age_num, y = cups_num, color = gender))+
  geom_jitter()+ #spreads the stacked points around
  theme_minimal()
```
Of course, you can stack it:
```{r}
ggplot(gact %>% filter(gender %in% c("Female", "Male", "Non-binary")), 
       #select gender identities with enough observations
       aes(x = age_num, y = cups_num, color = gender))+
  geom_jitter(alpha = 0.2)+ #alpha makes points more opaque
  geom_smooth(method = "lm")+ #y~x fit
  scale_color_brewer(palette = "Set1")+ #different color pre-set
  theme_minimal()+
  labs(title = "Age and coffee consumptions by gender",
       subtitle = "Linear fit added",
       x = "Age (approximate)", y = "Cups (daily)")+
  theme(legend.position = "bottom")
```

A few notes on how to make a good plot:

- always label your axis;

- give all the necessary information to read the plot (like what are the units);

- axis should cover the full range of possible values;

- colors (if any) should be distinct;

- keep it simple and don't over complicate things (i.e, having six colors for categories is a bad idea);

- don't use pie charts.

P.S. you can also apply ggplot to data:
```{r}
gact %>% 
  filter(gender == "Female") %>% 
  ggplot(aes(x = age_num))+
  geom_histogram(bins = 5)+
  theme_bw()
```

# Web-scraping

Let's have a small intermezzo. Often you will find some table, text, map, or database online which you can use. Some websites provide API (which is a whole different topic), but others don't. And so you may just want to get that data, but copy-pasting it by hand is not always an option, and that's where web-scraping is useful.

Some basics: websites are built using HTML markup language. The code of the page is open ("right click -> inspect" in browser), you can download it and find the elements of interest.

```{r}
#load some additional packages
#install.packages("rvest")
library(rvest)

webpage <- read_html("https://www.presidency.ucsb.edu/statistics/elections/2024")
#if you are interested, look more into these objects
table_data <- html_table(html_nodes(webpage, "table"), fill = TRUE)
elect_24 <- as.data.frame(table_data[[1]])[14:69,1:11]
names(elect_24) <- c("state", "total_votes", 
               "harris_votes", "harris_votes_share", "harris_ev",  
               "trump_votes", "trump_votes_share", "trump_ev",
               "other_votes", "other_votes_share", "other_ev")
write_csv(elect_24, "election_2024.csv")
head(elect_24)
```

## Merging

We have the dataset with all zip codes in the US:
```{R}
zips <- read_csv("zip_code_database.csv")
#did I already tell you that R has some amazing built-in functions?
zips$state_name <- state.name[match(zips$state,state.abb)]
head(zips)

glimpse(zips)
```
Use `zip` variable to merge with the zip_codes dataset. Let's check what potentially matches (change eval to T if you will copying from this notebook):
```{r, eval = F}
table(gact$zip %in% zips$zip)
```
For now, we can work with proper zip codes from our data. 

However, for optional practice: note that there are some shorter unmatched zip codes. You can actually link most of them to states by looking at the zip-code prefix.