---
title: "Homework 1: Data Cleaning, Merging and Aggregation in R"
author: "Emma Jay"
date: "January 24, 2025"
output: pdf_document
---
```{r setup}
#libraries
library(tidyverse)

#default theme for ggplot2
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 16))

#default parameters for knitr
knitr::opts_chunk$set(
  fig.width = 8,
  fig.asp = 0.618,
  fig.retina = 2,
  dpi = 150, 
  out.width = "70%"
)
```


## Part I: Survey Data Manipulation

### 1. Exploratory Data Analysis

```{r 1.1}
#load data
gactt_data <- read_csv("data/GACTT_RESULTS_ANONYMIZED_HW1.csv")

#print first 6 rows and last 4 rows
gactt_data %>%
  head(6)

gactt_data %>%
  tail(4)

#print class of each variable
gactt_data %>%
  sapply(class)

#plot - relationship between cups and party
gactt_data %>%
  ggplot(aes(x = party, fill = cups)) +
  geom_bar(position = "dodge") +
  labs(
    title = "Daily Coffee Consumption by Political Affiliation",
    x = "Party Affiliation",
    y = "Count", 
    fill = "Daily Cups of Coffee"
  )
```
__Graph Interpretation:__
The graph shows that Democrats were the most identified with group in the data set, followed by Republicans. In every political affiliation group, two cups of coffee was the most reported daily coffee intake, followed by one and then three (in every category but NA where NA was the third most responded). 

### 2. Merging Survey Data with ZIP Code Metadata

```{r 1.2}
#load zip code metadata
zip_codes <- read_csv("data/zip_code_database.csv")

#merge zip column with metadata to include state-level info in survey dataset
gactt_zip <- full_join(gactt_data, zip_codes, by = "zip")

#check match using number of NA values - should be same as unmatched zip

table(gactt_zip$zip %in% zip_codes$zip) %>%
  as.data.frame() %>%
  print()

table(is.na(gactt_zip$state)) %>%
  as.data.frame() %>%
  print()

```
__Number of Unmatched Zip Codes:__ 
There are 442 unmatched zip codes.

### 3. Aggregation and Insights
```{r 1.3}
#calculate average daily coffee consumption (cups) by state
avg_cups <- gactt_zip %>%
  mutate(cups = as.numeric(cups)) %>%
  group_by(state) %>%
  summarize(average_cups = mean(cups, na.rm = TRUE)) 
head(avg_cups)

#most preferred homebrew coffee method (home_brew) in each state
top_homebrew <- gactt_zip %>%
  filter(!is.na(home_brew)) %>%
  group_by(state, home_brew) %>%
  tally() %>%
  slice_max(order_by = n, n = 1) %>%
  ungroup() %>%
  select(state, home_brew)

head(top_homebrew)

#political affiliation breakdown (party) in each state (percentage of respondents identifying with democratic and republican)
politic_state <- gactt_zip %>%
  group_by(state, party) %>%
  count() %>%
  group_by(state) %>%
  mutate(percentage = n / sum(n) * 100) %>%
  ungroup () %>%
  select(state, party, percentage) 

head(politic_state)

#save aggregated data frame as a variable survey_state
survey_state <- avg_cups %>%
  left_join(top_homebrew, by = "state") %>%
  left_join(politic_state, by = "state")

head(survey_state)
```

## Part II: Election Data Analysis

### 1. Cleaning the Election Data
```{r 3.1}
#load 2024 election data
election_2024 <- read_csv("data/election_2024.csv")

#clean data

```

